---
title: "exercise-10"
format: html
editor: visual
---

## Exercise 10 - Practice ANOVA

```{r}
library(tidyverse)
library(ggplot2)
library(skimr)
library(stats)
library(broom)
library(infer)
library(effectsize)
library(car)

url <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv"
d = read_csv(url, col_names = TRUE)
```

> **Data source**:
>
> Tobias JA, et al. (2022). AVONET: Morphological, ecological and geographical data for all birds. *Ecology Letters* 25: 581--597.

-   Winnow the dataset to include only the following variables: **Species1**, **Family1**, **Order1**, **Beak.Length_Culmen**, **Beak.Width**, **Beak.Depth**, **Tarsus.Length**, **Wing.Length**, **Tail.Length**, **Mass**, **Habitat**, **Migration**, **Trophic.Level**, **Trophic.Niche**, **Min.Latitude**, **Max.Latitude**, **Centroid.Latitude**, **Primary.Lifestyle**, and **Range.Size**

-   Do a bit of exploratory data analysis with this dataset, e.g., using the {skimr} package. Which of the variables are categorical and which are numeric?

```{r}
avonet <- d |>
  select(
    Species1,
    Family1,
    Order1,
    Beak.Length_Culmen,
    Beak.Width,
    Beak.Depth,
    Tarsus.Length,
    Wing.Length,
    Tail.Length,
    Mass,
    Habitat,
    Migration,
    Trophic.Level,
    Trophic.Niche,
    Min.Latitude,
    Max.Latitude,
    Centroid.Latitude,
    Primary.Lifestyle,
    Range.Size
  )

head(avonet)
skimr::skim(avonet)
```

Upon utilizing the **skim()** function, we can analyze that the numeric variables are those which we can find summary statistics for that are output within the table

|     |                    |     |           |               |              |     |
|:----|:-------------------|----:|----------:|--------------:|-------------:|-----|
| 1   | Beak.Length_Culmen |   0 | 1.0000000 |  2.636346e+01 | 2.438601e+01 |     |
| 2   | Beak.Width         |   0 | 1.0000000 |  6.578545e+00 | 5.148234e+00 |     |
| 3   | Beak.Depth         |   0 | 1.0000000 |  8.060205e+00 | 7.585840e+00 |     |
| 4   | Tarsus.Length      |   0 | 1.0000000 |  2.873362e+01 | 2.484441e+01 |     |
| 5   | Wing.Length        |   0 | 1.0000000 |  1.247826e+02 | 9.343607e+01 |     |
| 6   | Tail.Length        |   0 | 1.0000000 |  8.665216e+01 | 6.108081e+01 |     |
| 7   | Mass               |   0 | 1.0000000 |  2.671472e+02 | 1.883035e+03 |     |
| 8   | Migration          |  23 | 0.9979108 |  1.288367e+00 | 6.202803e-01 |     |
| 9   | Min.Latitude       |  57 | 0.9948224 | -6.439430e+00 | 2.236939e+01 |     |
| 10  | Max.Latitude       |  57 | 0.9948224 |  1.150653e+01 | 2.331511e+01 |     |

### Challenge 1

#### One-Factor ANOVA and Interference

**Step 1**

-   Make boxplots of log(**Mass**) in relation to **Trophic.Level** and **Migration** behavior type. For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. Also, you will want to convert the variable **Migration** (which is scored as a number: "1", "2", or "3") from class numeric to either being classified as a factor or as a character (string) variable.

```{r}
avonet <- avonet |>
  mutate(Migration = as.factor(Migration)) |>
  filter(!is.na(Trophic.Level)) |>
  filter(!is.na(Migration)) |>
  filter(!is.na(Trophic.Niche))

ggplot(data = avonet, aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of log(Mass) by Trophic Level",
    x = "Trophic Level",
    y = "log(Mass)"
  )

ggplot(data = avonet, aes(x = Migration, y = log(Mass))) +
  geom_boxplot() +
  labs(
    title = "Boxplot of log(Mass) by Migration",
    x = "Migration",
    y = "log(Mass)"
  )
```

**Step 2**

-   Run linear models using the `lm()` function to look at the relationship between log(**Mass**) and **Trophic.Level** and between log(**Mass**) and **Migration**.

-   Examine the output of the resultant linear models. Is log(**Mass**) associated with either **Trophic.Level** or **Migration** category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

-   Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories.

```{r}
m1 <- lm(data = avonet, log(Mass) ~ Trophic.Level)
m2 <- lm(data = avonet, log(Mass) ~ Migration)

summary(m1)
summary(m2)

avonet <- avonet |>
  mutate(Migration = relevel(Migration, ref = "1"))

m3 <- lm(data = avonet |> filter(!is.na(Migration)), log(Mass) ~ Migration)
summary(m3)
```

The F-statistic of the **log(Mass) \~ Trophic.Level** and **log(Mass) \~ Migration** models both have

```         
p-value: < 2.2e-16
```

This indicates that there is a significant relationship when log(Mass) is regressed across different migratory categories and trophic levels. Furthermore, based on the associated p-values for the Migration 2 and Migration 3 categories, their appears to be a significant different when compared against the reference level (Migration 1) for differences in log(Mass).

**Step 3**

-   Conduct a post-hoc Turkey Honest Significant Differences test to also evaluate which **Migration** categories differ "significantly" from one another (see [**Module 20**](https://difiore.github.io/ada-2025/20-module.html))

```{r}
aov <- aov(data = avonet, log(Mass) ~ Migration)
posthoc <- TukeyHSD(aov, which = "Migration")
posthoc
```

Step 4

-   Use a permutation approach to inference to generate a null distribution of F statistic values for the model of log(**Mass**) in relation to **Trophic.Level** and calculate a p value for your original F statistic. You can do this either by programming your own permutation test (e.g., by shuffling values for the predictor or response variable among observations and calculating an F statistic for each replicate) or by using the {infer} workflow and setting `calculate(stat="F")`.

```{r}
avonet <- avonet |>
  mutate(logMass = log(Mass))

null_dist <- avonet |>
  specify(response = logMass, explanatory = Trophic.Level) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = 'permute') |>
  calculate(stat = "F")

obs_F <- avonet |>
  specify(response = logMass, explanatory = Trophic.Level) |>
  calculate(stat = "F")

p_val <- null_dist |>
  get_p_value(obs_stat = obs_F, direction = "greater")

obs_F
p_val
```

### Challenge 2

**Step 1**

-   Create the following two new variables and add them to AVONET dataset:

    -   **Relative beak length**, which you should calculate as the *residual* of log(**Beak.Length_Culmen**) on log(**Mass**).

    -   **Relative tarsus length**, which you should calculate as the *residual* of log(**Tarsus.Length**) on log(**Mass**).

```{r}
avonet_mod <- avonet |>
  mutate(Beak.Length_rel = residuals(lm(data = avonet, log(Mass) ~ log(Beak.Length_Culmen))),
         Tarsus.Length_rel = residuals(lm(data = avonet, log(Mass) ~ log(Tarsus.Length))))

# View(avonet_mod)
```

**Step 2**

-   Make a boxplot or violin plot of your new relative tarsus length variable in relation to **Primary.Lifestyle** and of your new relative beak length variable in relation to **Trophic.Niche**

```{r}
# Plot 1: Boxplot of Relative Tarsus Length by Primary Lifestyle
p_box_tarsus <- ggplot(data = avonet_mod , aes(x = Primary.Lifestyle, y = Tarsus.Length_rel)) +
  geom_boxplot() +
  labs( title = "Relative Tarsus Length by Primary Lifestyle",
        y = "Relative Tarsus Length (residuals)") +
  theme_minimal()

# Plot 2: Violin plot of Relative Tarsus Length by Primary Lifestyle
p_violin_tarsus <- ggplot(data = avonet_mod, aes(x = Primary.Lifestyle, y = Tarsus.Length_rel)) +
  geom_violin(trim = FALSE) +
  labs( title = "Relative Tarsus Length by Primary Lifestyle",
        y = "Relative Tarsus Length (residuals)") +
  theme_minimal()

# Plot 3: Boxplot of Relative Beak Length by Trophic Niche
p_box_beak <- ggplot(data = avonet_mod, aes(x = Trophic.Niche, y = Beak.Length_rel)) +
  geom_boxplot() +
  labs( title = "Relative Beak Length by Trophic Niche",
        y = "Relative Beak Length (residuals)") +
  theme_minimal()

# Plot 4: Violin plot of Relative Beak Length by Trophic Niche
p_violin_beak <- ggplot(data = avonet_mod, aes(x = Trophic.Niche, y = Beak.Length_rel)) +
  geom_violin(trim = FALSE) +
  labs( title = "Relative Beak Length by Trophic Niche",
        y = "Relative Beak Length (residuals)") +
  theme_minimal()

p_box_tarsus
p_violin_tarsus
p_box_beak
p_violin_beak
```

**Step 3**

-   Run ANOVA analyses to look at the association between geographic range size and the variable **Migration**. You should first drop those observations for which **Migration** is not scored and also look at the distribution of the variable **Range.Size** to decide whether and how it might need to be transformed. Based on the global model, is range size associated with form of migration? How much of the variance in your measure of range size is associated with **Migration** behavior style?

-   Given the regression coefficients returned in the output of the model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories. Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which **Migration** categories differ \"significantly\" from one another (see [**Module 20**](https://difiore.github.io/ada-2025/20-module.html)).

```{r}
# Analyzing Distribution of Range.Size
ggplot(data = avonet_mod, aes(x = Range.Size)) +
  geom_histogram(fill = "lightblue", color = "black") +
  labs(title = "Distribution of Range.Size")

avonet_mod <- avonet_mod |>
  mutate(Range.Size_log = log(Range.Size))

ggplot(data = avonet_mod, aes(x = Range.Size_log)) +
  geom_histogram(fill = "lightgreen", color = "black") +
  labs("Distribution of log(Range.Size")

```

The *log(Range.Size)* produces a **left-skewed normal distribution**

```{r}
# Linear model Testing the Log-transformed Range Size
m <- lm(data = avonet_mod, Range.Size_log ~ Migration)
m.aov <- Anova(m, formula = Range.Size_log ~ Migration)
m.aov

m.aov <- aov(data = avonet_mod, Range.Size_log ~ Migration)
TukeyHSD(m.aov)

```

```         
Residual standard error: 2.783 on 10929 degrees of freedom   
(49 observations effacées parce que manquantes) 
Multiple R-squared:  0.08681,	Adjusted R-squared:  0.08665  
F-statistic: 519.5 on 2 and 10929 DF,  
p-value: < 2.2e-16
```

 - **Global F-statistic** and its **p-value test** test whether *Migration* as a whole is associated with *log(Range.Size)*

 - The **adjusted R square** indicates the proportion of variance explained by *log(Range.Size).* In our model, our adjusted R square doesn't seem to be due to chance given the similarity to the **Multiple R-Squared**

For the associated regression summary, we find that there is **very low** explanation (**about** **8.7%**) of the proportion of model variance by the relationship between *Migration* and *log(Range.Size)*, but that there is a significant relationship between *Migration* and *log(Range.Size*) for the variance which the model does explain

**Step 4**

-   Winnow your original data to just consider birds from the Infraorder \"Passeriformes\" (song birds).

-   Run separate one-factor ANOVA analyses to look at the association between \[1\] relative beak length and **Primary.Lifestyle** and between \[2\] relative beak length and **Trophic.Level**. In doing so...

    -   Make boxplots of response variable by each predictor and by the combination of predictors.

    -   Run linear models for each predictor separately and interpret the model output.

```{r}
# Winnowing of avonet_mod into avonet_passeriformes
avonet_passeriformes <- avonet_mod |>
  filter(Order1 == "Passeriformes")

# View(avonet_passeriformes)


# Linear Model: Relative Beak Legnth and Primary Lifestyle 
m.passeriformes.1 <- lm(data = avonet_passeriformes, Beak.Length_rel ~ Primary.Lifestyle)
summary(m.passeriformes.1)

aov.passeriformes.1 <- Anova(m.passeriformes, formula = Beak.Length_rel ~ Primary.Lifestyle)
aov.passeriformes.1

ggplot(data = avonet_passeriformes, aes(x = Primary.Lifestyle, y = Beak.Length_rel)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Relative Beak Length by Primary Lifestyle",
       x = "Primary Lifestyle",
       y = "Relative Beak Length") +
  theme_minimal()

# Linear Model: Relative Beak Length by Trophic Level
m.passeriformes.2 <- lm(data = avonet_passeriformes, Beak.Length_rel ~ Trophic.Level)
summary(m.passeriformes.2)

aov.passeriformes.2 <- Anova(m.passeriformes, formula = Beak.Length_rel ~ Trophic.Level)
aov.passeriformes.2

ggplot(data = avonet_passeriformes, aes(x = Trophic.Level, y = Beak.Length_rel)) +
  geom_boxplot(fill = "lightpink") +
  labs(title = "Relative Beak Length by Trophic Level",
       x = "Trophic Level",
       y = "Relative Beak Length") +
  theme_minimal()

# Combination Plots: Relative Beak Length by the Combination of Primary Lifestyle and Trophic Level

ggplot(data = avonet_passeriformes, aes(x = interaction(Primary.Lifestyle, Trophic.Level, sep = " - "), y = Beak.Length_rel)) +
  geom_boxplot(fill = "lavender") +
  labs(title = "Relative Relative Beak Length by Primary Lifestyle and Trophic Level",
x = "Primary Lifestyle - Trophic Level",
y = "Relative Beak Length") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

```

**Step 5**

-   Run a two-factor model to look at the association between relative beak length and both **Primary.Lifestyle** and **Trophic.Level** among the passeriforms. Based on the model output, what would you conclude about how relative beak length is related to these two variables?

```{r}
lm_two_factor <- lm(data = avonet_passeriformes, Beak.Length_rel ~ Primary.Lifestyle + Trophic.Level)
summary(lm_two_factor)
```

-    **Overall Model Significance**

    -   The global F-statistic is **highly significant, however the adjusted R-squared value is quite low** (**about 17%**) indicating that for the two-variable model, the variance explained by regression model is rather low so *Primary.Lifestyle* and *Trophic.Level* are poorly explained by the *relative Beak Length*.

**Step 6**

-   Finally, run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term. To do this, you should modify your model formula using the colon operator (`:`) to specify the interaction, e.g., relative beak length \~ **Primary.Lifestyle** + **Trophic.Level** + **Primary.Lifestyle:Trophic.Level**. Based on the model output, what would you now conclude about how relative beak length is related to these two variables?

```{r}
lm_two_factor.2 <- lm(data = avonet_passeriformes, Beak.Length_rel ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level)
summary(lm_two_factor.2)
```

-   Both *Primary.Lifestyle* and *Trophic.Level* are **significantly** **explained** by the model that includes interaction, however, there is a **low** **amount of variance** explained by the model **(about 17%**) even when including the interaction term. Based on the relationships present within the combined interacting-term summary, *Primary.Lifestyle* and *Trophic.Level* share a **very significant relationship ( p-value \< 2e-16)** with the *relative beak length* in **all cases**. For the **interaction** term, only *Primary.LifestyleInsessorial:Trophic.LevelOmnivore* shows a significant relationship, with a p = 1.47e-0.5.

**Step 7**

-   Use the `interaction.plot()` function to visualize the interaction between **Primary.Lifestyle** and **Trophic.Level** (see [**Module 20**](https://difiore.github.io/ada-2025/20-module.html)).

```{r}
interaction.plot(
  x.factor = avonet_passeriformes$Primary.Lifestyle,
  trace.factor = avonet_passeriformes$Trophic.Level,
  response = avonet_passeriformes$Beak.Length_rel,
  fun = mean, 
  type = "b",
  leg
  col = c("red", "blue", "green"),
  pch = c(16,17,18),
  xlab = "Primary Lifestyle",
  ylab = "Mean Relative Beak Length",
  main = "Interaction: Primary Lifestyle by Trophic Level"
)
```

#### **Step 8**

In the exercise above, we really did not do any checking with this dataset to see if the data meet the primary assumptions for standard linear regression and ANOVA, which are that variables/residuals within each grouping level are roughly normally distributed and have roughly equal variances. Sample sizes within each grouping level should also be roughly equal. As noted in [**Module 20**](https://difiore.github.io/ada-2025/20-module.html), a general rule of thumb for \"equal\" variances is to compare the largest and smallest within-grouping level standard deviations and, if this value is less than 2, then it is often reasonable to presume the assumption may not be violated.

Use this approach to see whether variances in across groups in your various models (e.g., for **relative beak length \~ trophic level**) are roughly equal. Additionally, do a visual check of whether observations and model residuals within groups look to be normally distributed.

```{r}
# Distribution of Samples across Trophic Level
table(avonet_passeriformes$Trophic.Level)

# Compute Standard Deviation across Trophic Level
group_sd <- avonet_passeriformes |>
  group_by(Trophic.Level) |>
  summarise(sd_obs = sd(Beak.Length_rel, na.rm = TRUE),
            n = n())
group_sd

summary(lm(data = avonet_passeriformes, Beak.Length_rel ~ Trophic.Level))

avonet_passeriformes <- avonet_passeriformes |>
  mutate(resid_model = resid(lm(data = avonet_passeriformes, Beak.Length_rel ~ Trophic.Level)))
  

ratio_resid <- max(group_sd$sd_obs) / min(group_sd$sd_obs)
ratio_resid

```

Visually Inspection

```{r}
# Observed Values by Trophic Level
ggplot(avonet_passeriformes, aes(x = Beak.Length_rel)) +
  geom_histogram(fill = "lightblue", color = "black") +
  facet_wrap(~ Trophic.Level) +
  labs(title = "Histogram of Relative Beak Length by Trophic Level",
       x = "Relative Beak Length",
       y = "Frequency") +
  theme_minimal()

# Q-Q plots of residuals by Trophic.Level
ggplot(avonet_passeriformes, aes(sample = resid_model)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ Trophic.Level) +
  labs(title = "Q-Q Plot of Model Residuals by Trophic Level") +
  theme_minimal()

# Histograms of residuals by Trophic.Level
ggplot(avonet_passeriformes, aes(x = resid_model)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  facet_wrap(~ Trophic.Level) +
  labs(title = "Histogram of Model Residuals by Trophic Level",
       x = "Residuals",
       y = "Frequency") +
  theme_minimal()
```

Based on Q-Q plots, the data is mostly normally distributed with equivariance throughout its inner quartiles, however, the outer quartiles are more disorderly, especially in the first quartile which contains much less variance. This aligns with a cursory visual inspection, which indicates that the graphs are slightly left skewed.
